Developing a Visual Studio Code Language Support Extension for the Snail Programming Language

### 2023-06-30 paper.tex
feat: final rough draft (w/o conclusions)

Using documentation formed a huge role in the development of the Snail Language Support extension. We used the VS Code documentation for starter guides on how to build VS Code extensions. The documentation even had numerous sample extensions for various general exstension purposes that were useful to build on top of. Documentation also contained granular details on various options and specific functionality of the extension API. Similar documentation also exists for the language server protocol (LSP) and debug adapter protocol (DAP). Some of this documentation is hosted on VS Code's documentation website, but most is hosted on a separate website.

During development, Snail Language Support developers also made use of git documentation to look up specific syntax of commands. Some of this documentation is available while using git command line tools by passing the \lstinline{--help} flag.

Snail Language Support also creates documentation for the extension. One of the central goals of Snail Language Support development was to prioritize creating documentation during development. To do this, Snail Language Support leverages a few different documentation tools: Markdown, mermaidJS, and drawio.

Markdown is a plain text syntax that takes lightly marked-up text and converts it to clean HTML \cite{Gruber_2004}. The syntax is designed to be lightweight, allowing text files written with Markdown syntax to still be fully legible without converting to HTML. Since it's birth in 2004, Markdown has seen widespread, mainstream adoption across the industry. Popular tools like GitHub and VS Code offer built-in support for Markdown syntax \cite{GitHub_2023, Microsoft_2023k}. Markdown syntax has also strongly influenced the R Markdown syntax for R programming \cite{RStudio_2020}.

Snail Language Support uses Markdown to create a \lstinline{README.md} which describes valuable information about the extension for users and developers. Snail Language Support developers also used Markdown files as notepads to write down thoughts and discoveries during development. The hope in keeping these files around is that they may provide some valuable insight during future development.

In presenting the Snail Language Support extension (in posters, presentations, and this paper), we found the Mermaid diagramming tool quite helpful. Mermaid is a tool that allows developers to create high quality diagrams using Markdown-inspired text \cite{MermaidJS_2023a}. Mermaid diagrams can be developed in Mermaid's online editor, using third-party plug ins, or by using Mermaid's API in JavaScript files \cite{MermaidJS_2023b}.

@@

We also made use of a web-based diagramming tool called draw.io. Draw.io is a free, open source diagramming tool that runs in a web browser \cite{drawio_2023}. Draw.io provides a simple and easy-to-use user interface to allow developers to create high quality, customisable diagrams quickly. We use the third-party Draw.io Integration extension for VS Code to create and export draw.io diagrams into \lstinline{.png} files \cite{Dieterichs_2023}. This allows us to easily create diagrams such as those seen in figure \ref{fig:directory-structure} and figure \ref{fig:lsp}.

Despite software documentation being one of the central goals of Snail Language Support, we found it difficult to generate high quality documentation. First, it was difficult to motivate ourselves to generate documentation during development. Instead of documenting steps taken while troubleshooting, we found it easier to save time in the short term to continue developing the extension. Ultimately, this would result in longer development time as we would have to repeat unsuccessful troubleshooting steps when encountering similar issues in the future.

We also didn't realize how difficult making good documentation actually is. As frequent users of documentation, we know what lackluster documentation looks like when we see it. However, when trying to make documentation for Snail Language Support, we didn't realize the attention to detail and thoroughness that is required to generate high quality documentation for a piece of software. After developing Snail Language Support, we have a newfound appreciation for high quality documentation.

>>>>

Using documentation formed a huge role in the development of the Snail Language Support extension. We used the VS Code documentation for starter guides on how to build extensions, as well as to look at a few pre-built, example extensions. In addition to big-picture ideas, documentation also contained useful specific details on how to use the VS Code extension API. We used similar documentation for the language server protocol (LSP) and debug adapter protocol (DAP). Some of the LSP and DAP documentation was hosted on VS Code's documentation website, but most is hosted on separate, LSP and DAP specific websites.

During development, We also made use of git documentation to look up specific command syntax. Some of this documentation is available while using git command line tools by passing the \lstinline{--help} flag.

During development, we also wanted to focus on creating documentation for the Snail Language Support extension. To do this, Snail Language Support leverages a few different documentation tools: Markdown, Mermaid, and draw.io.

Markdown is a plain-text syntax that takes lightly marked-up text and converts it to clean HTML \cite{Gruber_2004}. The syntax is designed to be lightweight, allowing Markdown text files to be fully legible without converting to HTML. Since it's birth in 2004, Markdown has seen widespread, mainstream adoption across the software industry. Popular tools like GitHub and VS Code offer built-in support for Markdown syntax \cite{GitHub_2023, Microsoft_2023k}. Markdown syntax has also strongly influenced the R Markdown syntax for R programming \cite{RStudio_2020}. Markdown files end with the \lstinline{.md} file extension.

Snail Language Support uses Markdown to create a \lstinline{README.md} file which describes valuable information about the extension for users and developers. Snail Language Support developers also used Markdown files as notepads to write down thoughts and discoveries during development. These journals helped take note of significant discoveries and learnings during the development process. While these journals are far from polished, most are still stored in the Snail Language Support git repository, in the hopes they may prove useful during future development.

In presenting the Snail Language Support extension (in posters, presentations, and this paper), we found the Mermaid diagramming tool quite helpful. Mermaid is an open source tool that allows developers to create high quality diagrams using Markdown-inspired syntax \cite{MermaidJS_2023a}. Mermaid diagrams can be developed in Mermaid's online editor, using third-party plug-ins, or by using Mermaid's API in JavaScript files \cite{MermaidJS_2023b}.

@@

We also made use of a web-based diagramming tool called draw.io. Draw.io is a free, open source diagramming tool that runs in the web browser \cite{drawio_2023}. Draw.io provides a simple and easy-to-use user interface to allow developers to create high quality, customisable diagrams quickly. We use the third-party Draw.io Integration extension for VS Code to create and export draw.io diagrams into \lstinline{.png} files \cite{Dieterichs_2023}. This allows us to easily create free-form diagrams such as those seen in figure \ref{fig:directory-structure} and figure \ref{fig:lsp}.

As previously mentioned, creating software documentation \emph{throughout} the Snail Language Support development process was a central goal of this project. Despite this, we found it incredibly difficult to do this in practice. First, it was difficult to motivate ourselves set aside time to create documentation. Instead of documenting steps taken while troubleshooting, we found it easier to save time in the short term, skip documenting troubleshooting steps, and rush to further development. Unfortunately, more often than not we would encounter a similar issue later during development, and would have to repeat the same, painstakingly slow troubleshooting process. By trying to save time avoiding writing documentation, we ended up losing development time in the long run.

We also didn't realize how difficult it is to make high quality documentation. As frequent users of documentation, we know what lackluster documentation looks like when we see it. However, when trying to make documentation for Snail Language Support, we didn't realize the attention to detail and thoroughness that is required to generate high quality documentation for a piece of software. We found ourselves getting tired of creating reproducible installation instructions, linking relevant help sites, and trying to ensure instructions were not operating-system specific. After developing Snail Language Support, we have a newfound appreciation for high quality documentation.
